Large squeezing in multicomponent trapped Bose-Einstein condensates

### 2013-06-11 large_number_squeezing_with_atoms.tex
Minor changes to bog section.
Move the labels to the correct place in an extended eqnarray block

This system can be solved analytically, and expressions for quadrature, absolute and relative number squeezing can be derived. We present the solutions here; the full derivation can be found in \ref{appendixTwoModeDerivation}.
If we define
\begin{eqnarray}
s &=& \sin(\theta) \label{eqsDef} \\
c &=& \cos(\theta) \\
\theta &=& \Omega(t_3 - t_2) \\
\lambda_{ij} &=& \chi_{ij} \tau_{\mathrm{hold}} \\
A &=& \sqrt{n_a} \exp [n_a (e^{-i(\lambda_{aa}-\lambda_{ab})} - 1 )] \\
A_2 &=& n_a \exp [n_a (e^{-2i(\lambda_{aa}-\lambda_{ab})} - 1 )] \\
B &=& -i e^{i \phi}\sqrt{n_b} \exp [n_b (e^{-i(\lambda_{bb}-\lambda_{ab})} - 1 )] \\
B_2 &=& -n_b e^{2i\phi} \exp [n_b (e^{-2i(\lambda_{bb}-\lambda_{ab})} - 1 )] \\
D &=& AB^* - A^*B \label{eqDDef}
\end{eqnarray}
then the expectation values for number, number variance, and number difference variance after applying the final coupling field are given by
\begin{eqnarray}
N_a(t_3) &=& n_a c^2 + n_b s^2 + i c s D \\
%
N_b(t_3) &=& n_a s^2 + n_b c^2 - i c s D \\
%
{\mathrm{Var}} [ N_a](t_3) &=& n_a c^2 + n_b s^2 + 2 n_a n_b c^2 s^2 \\
       && + c^2 s^2 (D^2 - e^{i(\lambda_{aa} - \lambda_{bb})} B_2 A_2^* - e^{-i(\lambda_{aa} - \lambda_{bb})} B_2^* A_2) \nonumber \\
       && + i c s D (1-2 n_a c^2 -2 n_b s^2)  \nonumber\\
       && + 2 i c^3 s n_a (e^{-i(\lambda_{aa} - \lambda_{ab})} A B^* - e^{i(\lambda_{aa} - \lambda_{ab})} A^* B ) \nonumber \\
       && + 2 i c s^3 n_b (e^{i(\lambda_{bb} - \lambda_{ab})} A B^* - e^{-i(\lambda_{bb} - \lambda_{ab})} A^* B ) \nonumber\\
%
{\mathrm{Var}} [ N_b](t_3) &=&  n_a s^2 + n_b c^2 + 2 n_a n_b c^2 s^2 \nonumber \\
       && + c^2 s^2 (D^2 - e^{i(\lambda_{aa} - \lambda_{bb})} B_2 A_2^* - e^{-i(\lambda_{aa} - \lambda_{bb})} B_2^* A_2 ) \nonumber \\
       && + i c s D (-1+2 n_a s^2 + 2 n_b c^2)  \nonumber \\
       && + 2 i c^3 s n_b (e^{-i(\lambda_{bb} - \lambda_{ab})} A^* B - e^{i(\lambda_{bb} - \lambda_{ab})} A B^* ) \nonumber \\
       && + 2 i c s^3 n_a (e^{i(\lambda_{aa} - \lambda_{ab})} A^* B - e^{-i(\lambda_{aa} - \lambda_{ab})} A B^* ) \\
%
{\mathrm{Var}} [ N_a - N_b](t_3) &=& n_a + n_b + 4 i c s D (n_a - n_b)(s^2 - c^2) \nonumber \\
       && + 4 c^2 s^2 (2 n_a n_b +D^2 - e^{i(\lambda_{aa} - \lambda_{bb})} A_2^* B_2 - e^{-i(\lambda_{aa} - \lambda_{bb})} A_2 B_2^* ) \nonumber \\
       && + 4 i c s (c^2 - s^2) \left( n_a e^{-i(\lambda_{aa} - \lambda_{ab})} A B^* - n_a e^{i(\lambda_{aa} - \lambda_{ab})} A^* B \right. \nonumber \\
       && \,\,\,\,\, \left. + n_b e^{-i(\lambda_{bb} - \lambda_{ab})} A^* B - n_b e^{i(\lambda_{bb} - \lambda_{ab})} A B^* \right) \label{eqNumDiffVariance}
\end{eqnarray}
In the case where all the nonlinearities are equal, such that $\chi_{aa} = \chi_{ab} = \chi_{bb}$, the number variances all reduce to those of a coherent state, e.g. ${\mathrm{Var}}[N_a]=N_a$, and no number squeezing is possible. It should also be noted that, although there are three independent nonlinearities $\chi_{ij}$ in the Hamiltonian (\ref{eqTwoModeHamiltonian}), the expressions for the number variances only depend on differences. This means that number squeezing is parameterised by only two independent quantities, for example $\chi_{aa}-\chi_{ab}$ and $\chi_{aa}-\chi_{bb}$.

@@

[Insert motivation that the impact of Bogoliubov modes is detrimental here.  I could wave my hands and say that more modes is bad, as seen in the previous parts of the article, or I could refer to the squeezing of the Bogoliubov modes:
\begin{align}
  S_{a,\mathbf{k}\neq 0}(t_3) &= \frac{\frac{1}{2}\sum_{\mathbf{k}\neq0}n_a(\mathbf{k},t_3)\left[3 + \cos\theta + n_a(\mathbf{k}, t_3)\right]}{\sum_{\mathbf{k}\neq0} n_a(\mathbf{k}, t_3)},
\end{align}
which is always greater than 1, but this isn't convincing either as the total squeezing includes the number variance of the $\mathbf{k=0}$ mode as well as the correlations between the $\mathbf{k}=0$ mode and the $\mathbf{k}\neq 0$ modes.  If you do it naively, you find that when the Bog modes have a significant impact on the squeezing, they have also significantly changed it at $t=t_2$, which must be the same as $t=t_1$.  The problem is that the Bogoliubov theory used isn't number-conserving.  The best argument I have at the moment for Bog theory being a good indicator for when things go bad is that a significant Bog fraction implies a depleted condensate, and that will almost certainly make squeezing bad as it will invalidate the single-mode theory.  More freedom == bad?  In this case, the best evidence for the utility of Bog theory would seem to be Figure~\ref{fig1D_squeezing_and_bog_scaling}.]

Good squeezing requires that the Bogoliubov fraction or condensate depletion be small.  To better understand the scaling of $N_{\mathbf{k}\neq0}/N$, we will approximate the sum in \eqref{eqCondensateDepletion} with an integral.  In 1D \eqref{eqCondensateDepletion} becomes
\begin{align}
  \frac{N_{\mathbf{k}\neq 0}(t)}{N} &= \frac{(n_a \lambda_{aa} \cos\theta)^2}{N} \sum_{k \neq 0} \sinc^2\left(\sqrt{\omega^0_{k}(\omega^0_{k}+2 \chi_{aa}n_a)}\tau_\text{hold}\right).
\end{align}
In the limit that $\omega_{k_\text{min}} \tau_\text{hold} \ll 1$, where $k_\text{min} = 2\pi/L$ this discrete sum may be approximated by the integral
\begin{align}
  \frac{N_{\mathbf{k}\neq 0}(t)}{N} &\approx \frac{(n_a \lambda_{aa} \cos\theta)^2}{N} \int_{-\infty}^{+\infty} \sinc^2 \left(\sqrt{\omega^0_{k}(\omega^0_{k}+2 \chi_{aa}n_a)}\tau_\text{hold}\right) \frac{d k}{\Delta k},
\end{align}
where $\Delta k = 2\pi/L$.  Changing integration variables to $\Phi = \omega^{0}_{\mathbf{k}}\tau_\text{hold}$, the integral can be written as
\begin{align}
  \frac{N_{\mathbf{k}\neq 0}(t)}{N} &\approx \frac{(n_a \lambda_{aa} \cos\theta)^2}{N} \frac{L}{\sqrt{2}\pi\sqrt{\frac{\hbar}{M}\tau_\text{hold}}} \int_0^{\infty} \frac{1}{\sqrt{\Phi}}\sinc^2\sqrt{\Phi(\Phi + 2 \lambda_{aa} n_a)}\, d\Phi
\end{align}
Define $\Lambda_{a} = \lambda_{aa} n_a$ to simplify the expression and we get
\begin{align}
  \frac{N^\text{1D}_{\mathbf{k}\neq 0}(t)}{N} &\approx \frac{1}{N} \frac{\Lambda_a^2 L \cos^2\theta } {\sqrt{2}\pi\sqrt{\frac{\hbar}{M}\tau_\text{hold}}} f_\text{1D}(\Lambda_a),
\end{align}
where
\begin{align}
  f_\text{1D}(\Lambda) &= \int_0^{\infty} \frac{1}{\sqrt{\Phi}}\sinc^2\sqrt{\Phi(\Phi + 2 \Lambda)}\, d\Phi
\end{align}

>>>>

This system can be solved analytically, and expressions for quadrature, absolute and relative number squeezing can be derived. We present the solutions here; the full derivation can be found in \ref{appendixTwoModeDerivation}.
If we define
\begin{eqnarray}
s &=& \sin(\theta) \label{eqsDef} \\
c &=& \cos(\theta) \\
\theta &=& \Omega(t_3 - t_2) \\
\lambda_{ij} &=& \chi_{ij} \tau_{\mathrm{hold}} \\
A &=& \sqrt{n_a} \exp [n_a (e^{-i(\lambda_{aa}-\lambda_{ab})} - 1 )] \\
A_2 &=& n_a \exp [n_a (e^{-2i(\lambda_{aa}-\lambda_{ab})} - 1 )] \\
B &=& -i e^{i \phi}\sqrt{n_b} \exp [n_b (e^{-i(\lambda_{bb}-\lambda_{ab})} - 1 )] \\
B_2 &=& -n_b e^{2i\phi} \exp [n_b (e^{-2i(\lambda_{bb}-\lambda_{ab})} - 1 )] \\
D &=& AB^* - A^*B \label{eqDDef}
\end{eqnarray}
then the expectation values for number, number variance, and number difference variance after applying the final coupling field are given by
\begin{eqnarray}
N_a(t_3) &=& n_a c^2 + n_b s^2 + i c s D \\
%
N_b(t_3) &=& n_a s^2 + n_b c^2 - i c s D \\
%
{\mathrm{Var}} [ N_a](t_3) &=& n_a c^2 + n_b s^2 + 2 n_a n_b c^2 s^2 \nonumber\\
       && + c^2 s^2 (D^2 - e^{i(\lambda_{aa} - \lambda_{bb})} B_2 A_2^* - e^{-i(\lambda_{aa} - \lambda_{bb})} B_2^* A_2) \nonumber \\
       && + i c s D (1-2 n_a c^2 -2 n_b s^2)  \nonumber\\
       && + 2 i c^3 s n_a (e^{-i(\lambda_{aa} - \lambda_{ab})} A B^* - e^{i(\lambda_{aa} - \lambda_{ab})} A^* B ) \nonumber \\
       && + 2 i c s^3 n_b (e^{i(\lambda_{bb} - \lambda_{ab})} A B^* - e^{-i(\lambda_{bb} - \lambda_{ab})} A^* B ) \\
%
{\mathrm{Var}} [ N_b](t_3) &=&  n_a s^2 + n_b c^2 + 2 n_a n_b c^2 s^2 \nonumber \\
       && + c^2 s^2 (D^2 - e^{i(\lambda_{aa} - \lambda_{bb})} B_2 A_2^* - e^{-i(\lambda_{aa} - \lambda_{bb})} B_2^* A_2 ) \nonumber \\
       && + i c s D (-1+2 n_a s^2 + 2 n_b c^2)  \nonumber \\
       && + 2 i c^3 s n_b (e^{-i(\lambda_{bb} - \lambda_{ab})} A^* B - e^{i(\lambda_{bb} - \lambda_{ab})} A B^* ) \nonumber \\
       && + 2 i c s^3 n_a (e^{i(\lambda_{aa} - \lambda_{ab})} A^* B - e^{-i(\lambda_{aa} - \lambda_{ab})} A B^* ) \\
%
{\mathrm{Var}} [ N_a - N_b](t_3) &=& n_a + n_b + 4 i c s D (n_a - n_b)(s^2 - c^2) \nonumber \\
       && + 4 c^2 s^2 (2 n_a n_b +D^2 - e^{i(\lambda_{aa} - \lambda_{bb})} A_2^* B_2 - e^{-i(\lambda_{aa} - \lambda_{bb})} A_2 B_2^* ) \nonumber \\
       && + 4 i c s (c^2 - s^2) \left( n_a e^{-i(\lambda_{aa} - \lambda_{ab})} A B^* - n_a e^{i(\lambda_{aa} - \lambda_{ab})} A^* B \right. \nonumber \\
       && \,\,\,\,\, \left. + n_b e^{-i(\lambda_{bb} - \lambda_{ab})} A^* B - n_b e^{i(\lambda_{bb} - \lambda_{ab})} A B^* \right) \label{eqNumDiffVariance}
\end{eqnarray}
In the case where all the nonlinearities are equal, such that $\chi_{aa} = \chi_{ab} = \chi_{bb}$, the number variances all reduce to those of a coherent state, e.g. ${\mathrm{Var}}[N_a]=N_a$, and no number squeezing is possible. It should also be noted that, although there are three independent nonlinearities $\chi_{ij}$ in the Hamiltonian (\ref{eqTwoModeHamiltonian}), the expressions for the number variances only depend on differences. This means that number squeezing is parameterised by only two independent quantities, for example $\chi_{aa}-\chi_{ab}$ and $\chi_{aa}-\chi_{bb}$.

@@

[Insert motivation that the impact of Bogoliubov modes is detrimental here.  I could wave my hands and say that more modes is bad, as seen in the previous parts of the article, or I could refer to the squeezing of the Bogoliubov modes:
\begin{align}
  S_{a,\mathbf{k}\neq 0}(t_3) &= \frac{\frac{1}{2}\sum_{\mathbf{k}\neq0}n_a(\mathbf{k},t_3)\left[3 + \cos 2\theta + n_a(\mathbf{k}, t_3)\right]}{\sum_{\mathbf{k}\neq0} n_a(\mathbf{k}, t_3)},
\end{align}
which is always greater than 1, but this isn't convincing either as the total squeezing includes the number variance of the $\mathbf{k=0}$ mode as well as the correlations between the $\mathbf{k}=0$ mode and the $\mathbf{k}\neq 0$ modes.  If you do it naively, you find that when the Bog modes have a significant impact on the squeezing, they have also significantly changed it at $t=t_2$, which must be the same as $t=t_1$.  The problem is that the Bogoliubov theory used isn't number-conserving.  The best argument I have at the moment for Bog theory being a good indicator for when things go bad is that a significant Bog fraction implies a depleted condensate, and that will almost certainly make squeezing bad as it will invalidate the single-mode theory.  More freedom == bad?  In this case, the best evidence for the utility of Bog theory would seem to be Figure~\ref{fig1D_squeezing_and_bog_scaling}.]

Good squeezing requires that the Bogoliubov fraction or condensate depletion be small.  To better understand the scaling of $N_{\mathbf{k}\neq0}/N$, we will approximate the sum in \eqref{eqCondensateDepletion} with an integral.  In 1D \eqref{eqCondensateDepletion} becomes
\begin{align}
  \frac{N_{\mathbf{k}\neq 0}(t)}{N} &= \frac{(n_a \chi_{aa} \tau_\text{hold} \cos\theta)^2}{N} \sum_{k \neq 0} \sinc^2\left(\sqrt{\omega^0_{k}(\omega^0_{k}+2 \chi_{aa}n_a)}\tau_\text{hold}\right).
\end{align}
In the limit that $\omega_{k_\text{min}} \tau_\text{hold} \ll 1$, where $k_\text{min} = 2\pi/L$ this discrete sum may be approximated by the integral
\begin{align}
  \frac{N_{\mathbf{k}\neq 0}(t)}{N} &\approx \frac{(n_a \chi_{aa} \tau_\text{hold} \cos\theta)^2}{N} \int_{-\infty}^{+\infty} \sinc^2 \left(\sqrt{\omega^0_{k}(\omega^0_{k}+2 \chi_{aa}n_a)}\tau_\text{hold}\right) \frac{d k}{\Delta k},
\end{align}
where $\Delta k = 2\pi/L$.  Changing integration variables to $\Phi = \omega^{0}_{\mathbf{k}}\tau_\text{hold}$, the integral can be written as
\begin{align}
  \frac{N_{\mathbf{k}\neq 0}(t)}{N} &\approx \frac{(n_a \chi_{aa} \tau_\text{hold} \cos\theta)^2}{N} \frac{L}{\sqrt{2}\pi\sqrt{\frac{\hbar}{M}\tau_\text{hold}}} \int_0^{\infty} \frac{1}{\sqrt{\Phi}}\sinc^2\sqrt{\Phi(\Phi + 2 \lambda_{aa} n_a)}\, d\Phi
\end{align}
Define $\Lambda_{a} = n_a \chi_{aa} \tau_\text{hold}$ to simplify the expression and we get
\begin{align}
  \frac{N^\text{1D}_{\mathbf{k}\neq 0}(t)}{N} &\approx \frac{1}{N} \frac{\Lambda_a^2 L \cos^2\theta } {\sqrt{2}\pi\sqrt{\frac{\hbar}{M}\tau_\text{hold}}} f_\text{1D}(\Lambda_a),
\end{align}
where
\begin{align}
  f_\text{1D}(\Lambda) &= \int_0^{\infty} \frac{1}{\sqrt{\Phi}}\sinc^2\sqrt{\Phi(\Phi + 2 \Lambda)}\, d\Phi
\end{align}
